Generalized spaces for constructive algebra

### 2019-09-19 slides-herrsching2019.tex
Beginning of the slides for Proof and Computation 2019

\usepackage{pgfpages}
\setbeameroption{show notes on second screen}
\setbeamertemplate{note page}[plain]
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}

@@

\title{A general Nullstellensatz for generalised spaces}
\author{Ingo Blechschmidt}
\date{April 11th, 2019}

@@

\newcommand{\mynav}[3]{%
  foo
}

@@

@@

  \bigskip
  \[ \xymatrix{
    & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
    \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
    \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
  } \]
  \bigskip

  \begin{tikzpicture}
    \def\R{8pt}
    \node (title) {\hil{A general Nullstellensatz for generalised spaces}};
    \begin{pgfonlayer}{background}
      \draw[decorate, very thick, draw=mypurple]
        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}

@@

  Category Theory 2019 in Edinburgh \\
  July 9th, 2019
  \bigskip

@@

  \jnote{1}{\textbf{This talk in a nutshell:}

  There is a certain ring, the \emph{generic ring}, which is special in that
  it is \emph{conservative}: It has exactly those properties which are shared
  by any ring whatsoever. One could argue that it is this ring which we're
  implicitly referring to when we utter the phrase ``Let~$R$ be a ring.''.

  A surprising observation is that the generic ring is a field. This is not a
  contradiction because conservativity only pertains to properties which can be
  put in a certain syntactic form, dubbed \emph{geometric sequents}, and the
  field condition is not of that form.

  As a consequence, when verifying a geometric sequent for all rings, we may
  freely assume the field condition: Our proof will apply to the generic ring,
  because this ring happens to be a field, and its conclusion will spread to
  any ring whatsoever thanks to its syntactic form.

  An analogous story can be told for the generic group, the generic vector
  space and quite generally for the generic model of any \emph{geometric
  theory}. It is an interesting task to discover properties of such
  generic models. This talk presents a source of such properties, the
  \emph{topos-theoretic Nullstellensatz}, which is
  universal in the sense that any such property is a logical consequence of the
  Nullstellensatz.}
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight,trim=0 0 5.1cm 0,clip]{arthurs-seat-4}\end{minipage}}
\begin{frame}[plain]%
  \vspace*{\paperheight}\vspace*{-2.35cm}\hspace*{-1cm}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-7}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-0}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-5}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-9}%
  \jnote{1}{This talk owes a special debt to Alexander Oldenziel, who
  repeatedly suggested that the topic of this talk should be studied.

  The photos on the slide are of \emph{Arthur's Seat}, a beautiful extinct
  volcano in Edinburgh where parts of these slides have been created. It gives
  an excellent view of the city and is very much recommended for a hiking tour.
  }%
\end{frame}}
\addtocounter{framenumber}{-1}

\section{The generic model}

\subsection{The mystery of nongeometric sequents}

\begin{frame}{The mystery of nongeometric sequents}
  \jnote{1}{A \emph{geometric sequent} is a syntactic expression of the
  form~$(\varphi \seq{x_1\?X_1,\ldots,x_n\?X_n} \psi)$,
  where~$x_1\?X_1,\ldots,x_n\?X_n$ is a list of variable declarations,
  the~$X_i$ ranging over the available sorts, and~$\varphi$ and~$\psi$ are
  \emph{geometric formulas}. Such a sequent is read as ``in the
  context~$x_1\?X_1,\ldots,x_n\?X_n$, $\varphi$ entails $\psi$''.
  Geometric formulas are built from atomic propositions (equality and the
  available relation symbols) using the connectives~$\top$, $\bot$, $\wedge$, $\bigvee$
  (set-indexed disjunction) and~$\exists$. Geometric formulas may not
  contain~$\neg$, $\Rightarrow$, $\forall$.

  There is a notion of a \emph{model} of a geometric theory in
  any given topos. For instance, a ring in the usual sense is a model of the
  theory of rings in the topos~$\Set$. The structure sheaf~$\O_X$ of a scheme~$X$ is a
  model in the topos~$\Sh(X)$ of set-valued sheaves over~$X$.

  With \emph{topos} we mean Grothendieck topos, and as metatheory we use a
  constructive but impredicative flavour of English (which could be formalised
  by what is supported by the internal language of elementary toposes with an NNO).
  However the Nullstellensatz presented later makes no use of the
  subobject classifier, hence the results can likely be generalised to hold in
  a predicative metatheory or to hold for arithmetic universes.}

  \jnote{2-}{Among all models in any topos, the \emph{universal} or
  \emph{generic} one is special. It enjoys the universal property that any
  model in any topos can be obtained from it by pullback along an essentially
  unique geometric morphism. It is intriguing from a logical point of view
  because it has exactly those properties which are shared by any model in any
  topos.

  One could argue, with a certain amount of success, that the generic model of
  the theory of rings is what a mathematician implicitly refers to when she
  utters the phrase ``Let~$R$ be a ring''. This point of view is fundamental to
  the slogan \emph{continuity is geometricity}, as expounded for instance in
  \fixedhref{http://www.cs.bham.ac.uk/~sjv/GeoAspects.pdf}{Continuity and
  geometric logic} by Steve Vickers.}

  \jnote{3-}{Crucially, the conservativity statement only pertains to
  properties which can be put as geometric sequents. Generic models may have
  additional nongeometric properties. Because conservativity does not apply to
  them, they are not shared by all models in all toposes -- but any consequences
  which can be put as geometric sequents are.

  For instance, if we want to verify a geometric sequent for all local rings,
  we may freely use the displayed field axiom.
  Hence one reason why these nongeometric sequents are interesting is because
  they provide us with new reduction strategies (proofs by ``without loss of
  generality'').}

  Let~$\TT$ be a \explain{geometric theory}{1-}{red!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
    \small
    sorts,
    function symbols,
    relation symbols,
    geometric sequents as axioms
  \end{minipage}}, for instance the
  \explain{theory of rings}{1-}{blue!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
    \small
    \begin{tabbing}
      fun. symb.: \= \kill
      sorts: \> $R$ \\
      fun. symb.: \> $0$, $1$, $-$, $+$, $\cdot$ \\
      % rel. symb.: \> \emph{none} \\
      axioms: \> $(\top \vdash_{x,y:R} x y = y x)$, \ldots
    \end{tabbing}
  \end{minipage}}.

  \vspace*{1.8cm}

  \begin{center}
    \begin{tikzpicture}[ultra thick, node distance=7mm]
      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40] (a) {$\ZZ$};
      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of a] (b) {$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$};
      \node[regular polygon, regular polygon sides=5, draw=orange!80,
      fill=orange!20, right=of b, rounded corners=1pt, inner sep=0cm] (c) {$\O_X$};
      \node[star, rounded corners=1pt, star points=10, inner sep=0cm, draw=purple!80, fill=purple!20, right=of c] {$U_\TT$};
    \end{tikzpicture}
  \end{center}

  \vspace*{-0.5cm}
  \pause

  \justifying
  \textbf{Theorem.} There is a \hil{generic model}~$U_\TT$. It is
  \hil{conservative} in that
  for any \explainstub{geometric sequent}{yellow!70}~$\sigma$ the following notions coincide:
  \vspace*{-1.2em}
  \begin{enumerate}
    \item The sequent~$\sigma$ holds for~$U_\TT$.
    \item The sequent~$\sigma$ holds for any~$\TT$-model in any
    \explainstub{topos}{purple!20}.
    \item The sequent~$\sigma$ is provable modulo~$\TT$.
  \end{enumerate}

  \pause

  \justifying
  %\textbf{Main observation.} The generic model can validate nongeometric
  %sequents which are not provable modulo~$\TT$ and which do not hold for
  %arbitrary models in arbitrary toposes.

  \textbf{Observation (Kock).} The generic local ring is a field:
  \vspace*{-0.2cm}
  \[ (x = 0 \Rightarrow \bot) \vdash_{x:R} (\exists y\?R\_ xy = 1) \]
\end{frame}

\subsection{Construction of the generic model}

\begin{frame}{Construction of the generic model}
  The generic model is \hil{not} the same as \ldots
  \begin{itemize}
    \item the \hil{initial model} (think~$\ZZ$) or
    \item the \hil{free model on one generator} (think~$\ZZ[X]$).
  \end{itemize}
  Set-based models are \hil{too inflexible}.

  \textbf{Definition.} The \hil{syntactic site}~$\C_\TT$ has \ldots
  \begin{enumerate}
    \item objects: \explainstub{formulas-in-contexts}{red!20} $\{ x_1\?X_1, \ldots, x_n\?X_n\_ \varphi \}$
    \item morphisms: eqv. classes of \explainstub{provably functional
    formulas}{red!30}
    \item coverings: \explainstub{provably jointly surjective
    families}{purple!30}
  \end{enumerate}
  The topos of sheaves over~$\C_\TT$ is the \hil{classifying
  topos}~$\Set[\TT]$. The generic model interprets a sort~$X$ by the sheaf~$\yon\{x\?X\_\top\}$.

  \jnote{1}{In case the theory~$\TT$ is a Horn theory (for instance if it is
  an equational theory), the \emph{term algebra} (the set of terms in the
  empty context modulo provable equality) is a model of~$\TT$. While the
  term algebra does enjoy some nice categorical properties, it is in general
  \emph{not} the generic model.

  For instance, if~$\TT$ is the theory of rings, then the initial model
  is~$\ZZ$. This model validates some geometric sequents which are not
  validated by all rings, for instance~$(x^2 = 0 \seq{x:R} x = 0)$ or $(1 = 0
  \vdash \bot)$.

  In general, the generic model cannot be realised as a set-based model (with
  a set for each sort, a map for each function symbol and so on). Sets are too
  constant for this purpose; the flexibility of sheaves (``variable sets'') is
  required: The generic model lives in the topos of set-valued
  sheaves over~$\C_\TT$, called the \emph{classifying topos} of~$\TT$.}

  \jnote{2}{The notation in the definition of the objects of~$\C_\TT$ is meant
  to be suggestive, but should not be taken literally. The objects are
  syntactic expressions consisting of a context and a geometric formula in that
  context.

  A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
  equivalence class (modulo provable equivalence) of a geometric
  formula~$\theta$ such that~$\TT$ proves
  \begin{enumerate}
    \item ``$\theta$ is a relation on~$A \times B$'': $(\theta \seq{\vec x,
    \vec y} \varphi \wedge \psi)$
    \item ``$\theta$ is total'': $(\varphi \seq{\vec x} \exists \vec y\_
    \theta)$
    \item ``$\theta$ is single-valued'': $(\theta \wedge \theta[\vec y'/\vec y]
    \seq{\vec x,\vec y,\vec y'} \vec y = \vec y')$
  \end{enumerate}

  \vspace*{-0.6em}
  A family~$(\{ \vec x_i\_ \varphi_i \} \xrightarrow{[\theta_i]} \{ \vec y\_
  \psi \})_i$ is a covering iff~$\TT$ proves~$(\psi \seq{\vec y} \bigvee_i
  \exists \vec x_i\_ \theta_i)$.

  The slides experiment with using the symbol~``$\yon$'' for the Yoneda
  embedding~$\C_\TT \to \Set[\TT]$, as in \fixedhref{http://www.math.jhu.edu/~eriehl/elements.pdf}{Elements
  of~$(\infty,1)$-category theory} by Emily Riehl and Dominic Verity.}

  \jnote{3}{The special case that the generic model of a theory~$\TT$ can be realised as
  a model in~$\Set$ occurs iff~$\TT$ is Morita-equivalent to the
  empty theory, that is, iff~$\TT$ has exactly one model in any topos.

  The special case that there exists at least some conservative~$\TT$-model
  in~$\Set$ occurs iff~$\TT$ has a conservative geometric expansion to a theory
  which is Morita-equivalent to the empty theory.}
\end{frame}

\subsection{Going internal}

\begin{frame}{Working internally to toposes}
  \justifying
  Let~$\C$ be a site. We recursively define
  \[
    U \models \varphi \quad
    \text{(``$\varphi$ holds on~$U$'')}
  \]
  for objects~$U \in \C$ and \squiggly{formulas}~$\varphi$.
  Write~``$\Sh(\C) \models \varphi$'' for~$U \models \varphi$ for all~$U$.
  \only<1>{\footnotesize
  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
  U \models \top &\Ll& \text{true} \\
  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ \text{the
  empty family is a covering of~$U$} \\
  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
  U \models \varphi \wedge \psi &\Ll&
  \text{$U \models \varphi$ and $U \models \psi$} \\
  U \models \varphi \vee \psi &\Ll&
  \hcancel{\text{$U \models \varphi$ or $U \models
  \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $(U_i \to U)_i$} \\
  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
  U \models \varphi \Rightarrow \psi &\Ll&
  \text{for all~$V \to U$: }
  \text{$V \models \varphi$ implies $V \models \psi$} \\
  U \models \forall s \? F\_ \varphi(s) &\Ll&
  \text{for all $V \to U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
  % U \models \forall F\_ \varphi(F) &\Ll&
  % \text{for all $V \to U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
  U \models \exists s \? F\_ \varphi(s) &\Ll&
  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  &&
  \text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$}
  %U \models \exists F\_ \varphi(F) &\Ll&
  %\hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
  %\varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  %&&
  %\text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
  %&& \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
  \end{array} \]}

  \only<2>{
    \bigskip
    \begin{columns}[t]
      \begin{column}{0.45\textwidth}\justifying
        \textbf{Example.}
        $\Set \models \varphi$ if and only if $\varphi$ is true.
        \bigskip

        \textbf{Theorem.}
        If~$\Sh(\C) \models \varphi$ and if $\varphi$ intuitionistically
        entails~$\psi$, then also~$\Sh(\C) \models \psi$.
      \end{column}

      \begin{column}{0.45\textwidth}
        \href{https://arxiv.org/abs/1004.3802}{\includegraphics[width=\textwidth,align=t]{mike-shulman-stack-semantics}}
        \vspace*{1.5em}
      \end{column}
    \end{columns}
  }

  \jnote{1}{
    The internal language of a (Grothendieck or elementary) topos~$\E$ is a
    device which allows us to speak and reason about the objects and morphisms
    of~$\E$ in a naive element-based language close to the usual formal
    mathematical language. Using this language, objects of~$\E$ look like plain
    old sets [or types]; morphisms look like plain old maps between those sets;
    epimorphisms look like surjections; group objects look like groups; and so on.

    In particular, we can use the internal language to define what it means for a
    given~$\TT$-structure in~$\E$ to be a model -- namely iff it looks like a
    model from the internal point of view.

    The internal language can be implemented by the \emph{Kripke--Joyal semantics},
    a translation procedure which converts formulas of the internal language
    into external statements about the objects and morphisms of~$\E$. The slide
    displays some of the translation rules in the case that~$\E$ is a
    Grothendieck topos.
  }

  \jnote{2}{
    We can actually \emph{do mathematics} internally to a topos because the
    Kripke--Joyal semantics is sound with respect to intuitionistic logic:
    If~$\E \models \varphi$ and if~$\varphi$ intuitionistically entails a
    further formula~$\psi$, then~$\E \models \psi$.

    The Kripke--Joyal semantics can be extended to interpret unbounded
    quantification (``for all sets'' as opposed to ``for all elements of the
    particular set~$X$'') and dependent types. The former are for instance
    required to express universal properties (``for all groups'', ``for all
    rings''), and the latter are all over the place, even if their use might
    not be particularly highlighted.

    With these extensions, we can
    import all of everyday constructive impredicative mathematics into the
    internal world of a topos.

    Some illustrations of working with the internal language can be found in these
    sets of slides:
    \begin{itemize}
      \item
      \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-leipzig2018.pdf}{Slides
      for Jürgen Jost's group seminar at the MPI Leipzig}
      \item
      \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-como2018.pdf}{Slides
      for \emph{Toposes in Como}}
      (recording
      \fixedhref{https://www.youtube.com/watch?v=SuELfC5P_R8}{available})
    \end{itemize}
    A longer exposition, with pointers to the literature, can be found in Section~2 of
    \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these notes}.
  }
\end{frame}

\section[Nongeometric properties]{A selection of nongeometric properties}

\begin{frame}{A selection of nongeometric properties}
  The \hil{generic object} validates:
  \begin{enumerate}
    \item $\forall x,y\?U_\TT\_ \neg\neg(x = y)$.
    \item $\forall x_1,\ldots,x_n\?U_\TT\_ \neg \forall y\?U_\TT\_
    \bigvee_{i=1}^n y = x_i.$
    \item $(U_\TT)^{U_\TT} \cong 1 \amalg U_\TT$.
  \end{enumerate}

  The \hil{generic ring} validates:
  \begin{enumerate}
    \item $\forall x\?U_\TT\_ \neg\neg(x = 0)$.
    \item $\forall x\?U_\TT\_ (x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists
    y\?U_\TT\_ xy = 1)$.
  \end{enumerate}

  The \hil{generic local ring} validates:
  \begin{enumerate}
    \item $\neg \forall x\?U_\TT\_ \neg\neg(x = 0)$.
    \item $\forall a_0,\ldots,a_{n-1}\?U_\TT\_ \neg\neg \exists x\?U_\TT\_
      x^n + a_{n-1}x^{n-1} + \cdots + a_0 x^0 = 0$.
    \item \justifying Let~$\Delta = \{ \varepsilon \? U_\TT \,|\, \varepsilon^2 = 0 \}$.
    For any map~$f : \Delta \to U_\TT$, there are unique elements~$a,b \?
    U_\TT$ s.\,th.~$f(\varepsilon) = a + b\varepsilon$ for all~$\varepsilon
    \? \Delta$.
%   \item For any finitely presented~$U_\TT$-algebra~$A$,
%   the canonical map~$A \to (U_\TT)^{\Hom_{U_\TT}(A,U_\TT)}$
%   is an isomorphism.
  \end{enumerate}

  \jnote{1}{
    The generic object, the generic model of the theory which has exactly one
    sort and no function symbols, relations symbols or axioms, is
    slightly indecisive: On the one hand, up to a double negation, it is a
    subsingleton; on the other hand, it is infinite. This observation is due to
    Carsten Butz and Peter Johnstone
    (\fixedhref{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.16.4102&rep=rep1&type=pdf}{Classifying toposes for
    first-order theories}).

    The generic ring, the generic model of the theory of rings, is similarly
    indecisive. It is infinite in the following sense:
    \[ \forall x_1,\ldots,x_n \? U_\TT\_
      \bigl(\forall y \? U_\TT\_ \bigvee_{i=1}^n (y = x_i)\bigr) \Rightarrow
      1 = 0. \]

    The theory of local rings is the quotient theory of the theory of rings
    obtained by adding the axioms
    \[ (1 = 0 \vdash \bot) \ \text{and}\ ((\exists z\_
    (x+y)z = 1) \seq{x,y} (\exists z\_ xz = 1) \vee (\exists z\_ yz = 1)). \]
    (Assuming the axiom of choice, a ring is local in this sense
    iff it local in the usual sense (has exactly one maximal ideal).)

    The first displayed property of the generic local ring illustrates
    that nongeometric sequents need not be inherited by quotient theories.
  }

  \jnote{2}{
    All of the displayed properties give rise to reduction techniques:
    If we want to verify a geometric sequent for all rings, it suffices to
    verify it for the generic ring; but the generic ring has additional
    nongeometric properties not shared by every ring, such as the two displayed
    ones. (This was for instance used
    \fixedhref{https://www.sciencedirect.com/science/article/pii/0022404976900025}{by
    Anders Kock} and
    \fixedhref{https://link.springer.com/chapter/10.1007\%2FBFb0061835}{by
    Gonzalo Reyes}.)

    However, we face some challenges when pursuing these reduction techniques, including
    the following:
    \begin{enumerate}\justifying
      \item It is not easy to determine interesting and useful properties of
      the generic model.
      \item The set of validated nongeometric sequents changes slightly
      unpredictably when passing to quotient theories. For instance, when
      proving that a geometric sequent holds for all rings, we may assume that
      any element is \emph{not not} zero. But we may not assume this
      simplification if we want to verify a geometric sequent for all local
      rings (and if we want to exploit the given locality in the proof).
      \item There is only so much we want to state and prove in full generality
      for all rings, all local rings, all modules, and so on. We are often much
      more interested in properties of particular mathematical objects.
    \end{enumerate}
  }

  \jnote{3}{
    The firstly-mentioned problem on the previous page is alleviated by the
    Nullstellensatz presented in this talk, which gives a systematic and
    universal source of nongeometric sequents validated by the generic model.
    However, manual work is still required to reduce this set of sequents to a
    smaller, manageable one consisting of memorable properties while hopefully
    still preserving universality.

    To counter the third problem, it's prudent to consider geometric theories
    which depend on a given mathematical object of interest. For instance,
    given a ring~$A$, we can consider the theory of prime ideals of~$A$, of
    complemented prime ideals, of filters, and so on. The classifying toposes
    of these theories are of independent interest -- in fact they are sheaf
    toposes over certain important spaces in algebraic geometry -- and
    nongeometric sequents validated by their generic models bundle nontrivial
    information about~$A$. More details are on the following slide.
  }
\end{frame}

\begin{frame}{An application in commutative algebra}
  Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).

  The \hil{generic local localisation}~$A^\sim$ of~$A$ validates:

  \vspace*{-1.8em}
  \setbeamercolor{block body}{bg=red!30}
  \setbeamercolor{structure}{fg=purple}
  \begin{varblock}{\textwidth}{}
    $A^\sim$ is a \hil{field}: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_
    xy = 1) \Rightarrow x = 0)$.

    $A^\sim$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}:
    $\forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y$.

    \mbox{$A^\sim$ is \hil{anonymously Noetherian}.}\\[-1.2em]
  \end{varblock}

  This observation unlocks a short and conceptual proof of Grothendieck's
  generic freeness lemma in algebraic geometry.

  \jnote{1}{A \emph{local localisation} of a ring~$A$ is a
  localisation~$A[S^{-1}]$ which is a local ring. The classifying topos of
  local localisations of~$A$ is none other than the topos of sheaves over the
  spectrum of~$A$, as studied in algebraic geometry; and the generic model is
  also known as the structure sheaf~$\O_{\operatorname{Spec}(A)}$.

  Assuming the Boolean prime ideal theorem, the geometric sequents
  validated by~$A^\sim$ are easy to describe: They are precisely those which
  are validated by all the stalks~$A_\ppp$ of~$A$.

  But~$A^\sim$ enjoys further unique properties which are not shared by the
  stalks of~$A$, other localisations of~$A$, quotients of~$A$ or indeed any
  reasonable construction. Three of these are displayed on the slide.
  (A ring is \emph{anonymously Noetherian} iff each of its ideals is
  \emph{not not} finitely generated. Textbook proofs of Hilbert's basis theorem
  are constructively valid for this Noetherian condition.)

  The object~$A^\sim$ strikes a fine balance: On the one hand, it is still
  close to~$A$, so that information learned about~$A^\sim$ teaches us
  about~$A$; on the other hand, it enjoys unique properties rendering it
  simpler than~$A$.

  This balance allows for a simple and conceptually satisfying proof of
  \emph{Gro\-then\-dieck's generic freeness lemma}, an important theorem in
  algebraic geometry. Details can be found in
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-padova2018.pdf}{this
  set of slides}.}
\end{frame}

\section[A systematic source]{A systematic source of nongeometricity?}

\begin{frame}{A systematic source of nongeometricity?}
  \jnote{1-2}{
    Mimicking the synthetic approach to differential geometry, synthetic
    algebraic geometry is a framework for algebraic geometry in which schemes
    are modelled by plain old sets, morphisms of schemes by plain old maps
    between those sets, group schemes by plain old groups, and so on. Unlike
    its close cousin, it is far less developed; some first steps are outlined
    in Sections~19 and~20 of
    \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these notes}.

    Synthetic algebraic geometry is carried out internally to the \emph{big
    Zariski topos} of a given base scheme; in the special case that the base
    scheme is the terminal scheme~$\Spec(\ZZ)$, this topos is just the
    classifying topos of local rings. The relevant generic model living in the
    big Zariski topos is denoted~``$\affl$'' because it coincides with the
    functor of points of the affine line.
  }

  \jnote{2}{
    Marc Bezem, Ulrik Buchholtz and Thierry Coquand answered in their 2017 paper
    \fixedhref{https://arxiv.org/abs/1712.07743}{Syntactic forcing
    models for coherent logic} Gavin Wraith's question in the negative. (As
    Thierry remarked at 6WFTop, this is even if one takes care to phrase
    the question in a way to exclude the trivial counterexamples given by
    instances of the law of excluded middle in the language of~$\TT$.) If the
    answer had been positive, this would have given a neat, if somewhat hard to
    use in practice, characterisation of the formulas validated by the generic model.
  }

  \jnote{3}{
    The topos-theoretic \emph{Nullstellensatz}, to be presented on the next
    slides, answers the displayed three questions in the affirmative.

    Briefly, the Nullstellensatz is a certain statement in the language of a
    given geometric theory~$\TT$ which is
    \begin{itemize}\justifying
      \item validated by the generic~$\TT$-model,
      \item typically not validated by other~$\TT$-models, and
      \item such that any statement validated by the generic~$\TT$-model can be
      deduced, in intuitionistic logic, from the axioms of~$\TT$ and the
      Nullstellensatz.
    \end{itemize}

    We believe that this characterisation is as explicit as it can get, but
    would be delighted to be surprised by a future improvement. We stumbled on
    it by playing with the synthetic quasicoherence statement, not least thanks
    to encouragement by Alexander Oldenziel. However the route from that
    statement to the Nullstellensatz is not quite direct; it turns out that
    synthetic quasicoherence is a corollary of a specialisation of a
    higher-order version of the Nullstellensatz to Horn theories.
  }

  \justifying
  \textbf{Empirical fact.}
  In \hil{synthetic algebraic geometry}, every known property of~$\affl$ followed from
  its \hil{synthetic quasicoherence}:

  \begin{quote}
    \emph{\mbox{For any finitely presented~$\affl$-algebra~$A$, the ``double dual'' map}
    \[ A \longrightarrow (\affl)^{\Hom_\affl(A,\affl)},\ s \longmapsto (x
    \mapsto x(s)) \]
    is an isomorphism of~$\affl$-algebras.}
  \end{quote}

  \vspace*{1.0em}
  \begin{enumerate}
    \item Does a general metatheorem explain this observation?
    \item Is there a systematic source in any classifying topos?
    \item Is there even an exhaustive source?
  \end{enumerate}
  \vspace*{-3.3em}

  \visible<2>{\begin{tikzpicture}[overlay]
    \draw[fill=white, draw=white, opacity=0.95] (-1,1.4) rectangle (\paperwidth,5.4);
    \node[anchor=south west,inner sep=0] (image) at (0,0.9) {\vbox{
      \only<2>{
	\centering
	\includegraphics[width=0.9\textwidth]{wraith-some-recent-developments-in-topos-theory} \\
	\scriptsize
        Gavin Wraith. \emph{Some recent developments in topos theory.} \\ In:
        Proc.\@ of the ICM (Helsinki, 1978).
      }
    }};
  \end{tikzpicture}}%

  \[ \xymatrix{
    \text{$\TT$ proves~$\alpha$\;} \ar@/_1pc/@{=>}[r] &
    \text{\;$\alpha$ holds for~$U_\TT$\;} \ar@/_1pc/@{=>}[r]
    \ar@/_1pc/@{=>}@[red][l]
    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\!\texttimes\!}}}}|{} &
    \text{\;$\alpha$ is~$\TT$-redundant}
    \ar@/_1pc/@{=>}@[red][l]
    |{\SelectTips{cm}{}\textbf{\scalebox{3}{\textcolor{red}{\!\texttimes\!}}}}|{}
  } \]
\end{frame}

\section{A topos-theoretic Nullstellensatz}

\newcommand{\nullstellensatz}{%
  \justifying
  \textbf{Theorem.} Internally to~$\Set[\TT]$:

  \vspace*{-0.3em}
  \begin{quote}\emph{\mbox{For any \explainstub{geometric$^\star$ sequent~$\sigma$}{red!20} over
  the \explainstub{signature of~$\ul{\TT}/U_\TT$}{blue!20},}
  if~\explainstub{$\sigma$ holds for~$U_\TT$}{yellow!70}, then~\explainstub{$\ul{\TT}/U_\TT$
  proves~$\sigma$}{yellow!70}.}
  \end{quote}
  \vspace*{-0.6em}
}

\begin{frame}{A topos-theoretic Nullstellensatz}
  \jnote{1-2}{
    By~$\ul{\TT}$ we mean the geometric theory internal to~$\Set[\TT]$ obtained
    by pulling back the set of sorts of~$\TT$, the set of function symbols and so on
    along the unique geometric morphism~$\Set[\TT] \to \Set$. For instance,
    if~$\TT$ is the theory of rings, then from the internal point of view
    of~$\Set[\TT]$, the theory~$\ul{\TT}$ will again be the theory of rings.

    The theory~$\ul{\TT}/U_\TT$ will be defined on the next slide. It is a certain
    geometric theory internal to~$\Set[\TT]$.

>>>>

%\usepackage{pgfpages}
%\setbeameroption{show notes on second screen}
%\setbeamertemplate{note page}[plain]
%\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
\newcommand{\jnote}[2]{}

@@

\title{Generalized spaces for constructive algebra}
\author{Ingo Blechschmidt}
\date{September 2019}

@@

@@

\tikzstyle{topos} = [draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt, inner ysep=10pt]
\tikzstyle{title} = [fill=mypurple, text=white]

\input{images/primes.tex}
%\renewcommand{\sieve}[2]{SIEVE}
%\renewcommand{\fakesieve}[2]{SIEVE}

\newcommand{\drawbox}[4]{
  \node[topos, #4] [fit = #3] (#1) {};
  \node[title] at (#1.north) {#2};
}

\newcommand{\muchstuff}{
  \includegraphics[height=3em]{filmat}
  \scalebox{0.5}{\sieve{14}{2}}
}

\newcommand{\muchstuffplaceholder}{
  \includegraphics[height=3em]{filmat-placeholder}
  \scalebox{0.5}{\fakesieve{14}{2}}
}

\newcommand{\fewstuff}{
  \includegraphics[height=3em]{filmat}
  \scalebox{0.5}{\sieve{7}{2}}
}

\newcommand{\threeblobs}{
  \colorbox{mypurple}{\ \ }\quad
  \colorbox{mypurple}{\ \ }\quad
  \colorbox{mypurple}{\ \ }
}

@@

  \begin{tikzpicture}
    \def\R{8pt}
    \node (title) {\hil{\,Generalized spaces for constructive algebra\,}};
    \begin{pgfonlayer}{background}
      \draw[decorate, very thick, draw=mypurple]
        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}
  \bigskip

  \begin{columns}
    \small
    \begin{column}{0.35\textwidth}
      \centering
      \bignumber{1}

      \phantom{x}\hil{Locales}
      \medskip

      \includegraphics[height=4.5em]{phantoms.png}
    \end{column}
    \hspace*{0.5em}
    \begin{column}{0.28\textwidth}
      \centering
      \bignumber{2}

      \hil{Sheaf models}
      \medskip

      \begin{tikzpicture}
        \node (overview) {
          \scalebox{0.35}{\sieve{4}{3}}
        };
        \def\R{8pt}
        \begin{pgfonlayer}{background}
        \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
          ($(overview.south west) + (\R, 0)$) arc(270:180:\R) --
          ($(overview.north west) + (0, -\R)$) arc(180:90:\R) --
          ($(overview.north east) + (-\R, 0)$) arc(90:0:\R) --
          ($(overview.south east) + (0, \R)$) arc(0:-90:\R) --
          cycle;
        \end{pgfonlayer}
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.4\textwidth}
      \centering
      \bignumber{3}

      \hil{Constructive algebra}
      \medskip

      \includegraphics[height=4.5em]{calabi-yau}
    \end{column}
  \end{columns}
  \bigskip

@@

  Proof and Computation \\
  September 20th to 26th, 2019
  \bigskip

@@

\end{frame}}
